---
title: "Reproducible Research: Peer Assessment 1"
author: "Edgar Sánchez M."
date: "March 30, 2018"
output: 
  html_document:
    keep_md: true
---

## Prepare environment
```{r}
setwd("~/R/RepData_PeerAssessment1")
rm(list = ls())

unzip("activity.zip", overwrite = T)
```
```{r ldLibs, echo=FALSE}
library(ggplot2)
library(dplyr)
library(xtable)
```

## Loading and preprocessing the data

If the data.frame already exists then don't load it again
```{r}
if ( !(exists("activity"))) {
  activity0 <- read.csv("activity.csv")
}
activity0$date <- as.Date(activity0$date, "%Y-%m-%d")
```

Setting all the NA's to zero
```{r}
activity <- activity0
activity[is.na(activity$steps),]$steps <- 0
```


## What is mean total number of steps taken per day?

Calculate summary data.frame "**summary_sum**" by date
```{r}
activity_sum <- group_by(activity, date)  %>% 
  summarise(totalSteps=sum(steps), meanSteps=mean(steps), medianSteps=median(steps)) %>% 
  arrange(date)
activity_sum <- as.data.frame(activity_sum)
```

Create the histogram showing average steps taken by day
```{r avgPlot1, fig.width=10}
barplot(
  activity_sum$meanSteps, 
  names.arg = activity_sum$date, 
  las = 2, 
  cex.names=0.6, 
  xlab = "Day", 
  ylab = "Average Steps",
  main = "Average Steps by Day",
  col = "lightblue"
)
```

The mean and median steps per day is shown in the following table:

```{r tableMeans1, results="asis", echo=FALSE}
toShow <- select( activity_sum, date, meanSteps, medianSteps)
toShow$date <- as.character(toShow$date, "%Y-%m-%d")
colnames( toShow ) <- c( "Date", "Mean Steps", "Median Steps" )

print( 
      xtable(toShow), 
      type="html",
      include.rownames=FALSE,
      html.table.attributes='border="1" style="width: 60%; border: 3px solid black; border-collapse: collapse"'
    )
```

## What is the average daily activity pattern?

Summarize info by Interval, generate data.frame "**activity_sum1**"
```{r}
activity_sum1 <- group_by(activity, interval) %>% 
    summarise(totalSteps = sum(steps), avgSteps = mean(steps))
```

Determine which interval has the maximum values
```{r}
maxInterval <- as.numeric( activity_sum1[activity_sum1$totalSteps == max(activity_sum1$totalSteps), 1] )
```

Which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps? 

Answer:  Interval **#`r maxInterval`**

### Generate the time series plot
```{r timeSeries, fig.width=10}
plot( 
  activity_sum1$interval, 
  activity_sum1$avgSteps, 
  type = "l",
  xlab = "Interval",
  ylab = "Average steps",
  main = "Average steps by interval"
)
abline( v = maxInterval, col = "red")
text( maxInterval, 0 , paste("Max interval = ", maxInterval) , pos=4)
```

## Imputing missing values

Obtain a vector with NA's from the original data set:
```{r}
activityNAS <- activity0[ is.na( activity0$steps ), ] 
numOfNas <- nrow( activityNAS )
```

The number of NA's present in the original data set is: **`r numOfNas`**

We're gonna use the **average number of steps per interval** to assign it to the NA values. For this we use data.frame "*activity_sum1*" created before.
```{r}
activityNAS <- merge( activityNAS, activity_sum1, all = T ) %>% select( steps=avgSteps, date, interval )
activity1 <- rbind( activity0[ !is.na( activity0$steps ), ] , activityNAS )
```

Summarize the modified data.frame, the result is stored in data.frame "**activity_sum2**"
```{r}
activity_sum2 <- group_by(activity1, date)  %>% 
  summarise(totalSteps=sum(steps), meanSteps=mean(steps), medianSteps=median(steps), countSteps=n()) %>% 
  arrange(date)
activity_sum2 <- as.data.frame(activity_sum2)
```

Generate the new plot with the NA values replaced
```{r barPlot2, fig.width=10}
barplot(
  activity_sum2$meanSteps, 
  names.arg = activity_sum2$date, 
  las = 2, 
  cex.names=0.6, 
  xlab = "Day", 
  ylab = "Average Steps",
  main = "Average Steps by Day (with NAS)",
  col = "lightgreen"
)
```
The mean and median steps per day is shown in the following table:

```{r tableMeans2, results="asis", echo=FALSE}
toShow <- select( activity_sum2, date, meanSteps, medianSteps)
toShow$date <- as.character(toShow$date, "%Y-%m-%d")
colnames( toShow ) <- c( "Date", "Mean Steps", "Median Steps" )

print( 
      xtable(toShow), 
      type="html",
      include.rownames=FALSE,
      html.table.attributes='border="1" style="width: 60%; border: 3px solid black; border-collapse: collapse"'
    )
```

## Are there differences in activity patterns between weekdays and weekends?

In order to determine the label for Weekdays and Weekends, the following function is defined:
```{r}
isWeekend <- function( x ){
  if( as.POSIXlt(x)$wday %in% c(0,6)){
    "Weekend"
  } else {
    "Weekday"
  }
}
```
If posix weekday number is 0 (Sunday) or 6 (Saturday), then the label is "Weekend" else it's "Weekday"

Use the function to generate column labeling what type of day it is
```{r}
weekDayVect <- sapply( activity1$date, isWeekend )
activity1 <- cbind( activity1, dayType = weekDayVect)
```

Summarize the information by DayType and Interval
```{r}
plot_data <- aggregate(steps~ dayType + interval, data=activity1 , mean )
```

Generate plot, there is more step activity during the first intervals of Weekdays:
```{r plotWeekends, fig.width=10, fig.height=8}
daysType <- qplot( 
              interval, 
              steps, 
              data=plot_data, 
              facets=dayType~. , 
              geom = "line",
              xlab = "Interval",
              ylab = "Number of Steps"
            )
print(daysType)
```

